---
title: "Sushi Rankings Analysis"
subtitle: "Voting rules, agreement, and preference clustering"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
    df-print: kable
execute:
  warning: false
  message: false
  cache: false
---

```{r setup, include=FALSE}
library(BayesMallows)
library(tidyverse)
library(mclust)
library(knitr)

data("sushi_rankings", package = "BayesMallows")
sushi  <- sushi_rankings
items  <- colnames(sushi)
n_voters <- nrow(sushi)   # 5 000
n_items  <- ncol(sushi)   # 10

# ── Part 1: First preferences ────────────────────────────────────────────────
fp_counts <- colSums(sushi == 1)
fp_prop   <- fp_counts / n_voters
fp_df <- tibble(Item = items, Count = fp_counts, Proportion = fp_prop) |>
  arrange(desc(Count))
fp_winner <- fp_df$Item[1]

# ── Part 1: IRV ───────────────────────────────────────────────────────────────
run_irv <- function(mat) {
  active     <- colnames(mat)
  n          <- nrow(mat)
  rounds     <- list()
  eliminated <- character(0)

  repeat {
    top_choice <- apply(mat[, active, drop = FALSE], 1,
                        function(r) active[which.min(r)])
    counts <- setNames(integer(length(active)), active)
    tbl    <- table(top_choice)
    counts[names(tbl)] <- as.integer(tbl)
    rounds[[length(rounds) + 1]] <- counts

    if (max(counts) > n / 2 || length(active) == 1) {
      winner <- names(which.max(counts))
      break
    }

    loser      <- names(which.min(counts))
    eliminated <- c(eliminated, loser)
    active     <- active[active != loser]
  }
  list(winner = winner, rounds = rounds, eliminated = eliminated)
}

irv_res    <- run_irv(sushi)
irv_winner <- irv_res$winner

# ── Part 1: Condorcet ─────────────────────────────────────────────────────────
pw_wins <- matrix(0L, n_items, n_items, dimnames = list(items, items))
for (i in seq_len(n_items))
  for (j in seq_len(n_items))
    if (i != j)
      pw_wins[i, j] <- sum(sushi[, i] < sushi[, j])

condorcet_winner <- NA_character_
for (item in items) {
  others <- items[items != item]
  if (all(pw_wins[item, others] > n_voters / 2)) {
    condorcet_winner <- item
    break
  }
}

cycle_example <- NULL
if (is.na(condorcet_winner)) {
  beats <- pw_wins > n_voters / 2
  for (a in items) {
    for (b in items[items != a]) {
      if (beats[a, b]) {
        for (cc in items[items != a & items != b]) {
          if (beats[b, cc] && beats[cc, a]) {
            cycle_example <- c(a, b, cc); break
          }
        }
      }
      if (!is.null(cycle_example)) break
    }
    if (!is.null(cycle_example)) break
  }
}

condorcet_label <- if (!is.na(condorcet_winner)) condorcet_winner else "none (preference cycles exist)"

# ── Part 2: Kendall's W ───────────────────────────────────────────────────────
m   <- n_voters
n   <- n_items
R_j <- colSums(sushi)          # column rank sums
R_bar <- m * (n + 1) / 2      # expected column sum under randomness
S   <- sum((R_j - R_bar)^2)
W   <- 12 * S / (m^2 * (n^3 - n))
chi_sq_W  <- m * (n - 1) * W
df_W      <- n - 1
p_W       <- pchisq(chi_sq_W, df = df_W, lower.tail = FALSE)

W_interp <- case_when(
  W < 0.1  ~ "very weak",
  W < 0.3  ~ "weak",
  W < 0.5  ~ "moderate",
  W < 0.7  ~ "strong",
  TRUE     ~ "very strong"
)

# ── Part 3: Clustering (mclust) ───────────────────────────────────────────────
# Rankings sum to 55 per row → data lies on a 9-D hyperplane.
# Project onto first 9 PCs (removing the zero-variance 10th PC) so mclust
# does not encounter a singular covariance matrix.
pca_sushi  <- prcomp(sushi, center = TRUE, scale. = FALSE)
sushi_9pc  <- pca_sushi$x[, 1:9]

set.seed(42)
bic_obj <- mclustBIC(sushi_9pc, G = 1:4)

mc_fit     <- Mclust(sushi_9pc, G = 1:4, x = bic_obj)
best_G     <- mc_fit$G
best_model <- mc_fit$modelName

z_matrix   <- mc_fit$z                # posterior membership probabilities
cl_assign  <- mc_fit$classification   # hard assignment

cl_sizes <- tibble(
  Cluster = seq_len(best_G),
  Count   = as.integer(table(cl_assign)),
  Proportion = round(as.numeric(table(cl_assign)) / n_voters, 3)
)

# Average rank per cluster (lower = more preferred)
cl_avg_ranks <- lapply(seq_len(best_G), function(k) {
  colMeans(sushi[cl_assign == k, ])
})

# Top 3 most preferred items per cluster
top3_per_cluster <- map_chr(cl_avg_ranks, function(avg_r) {
  paste(names(sort(avg_r))[1:3], collapse = ", ")
})

# Plain-language cluster descriptions for summary
cl_desc <- map_chr(seq_len(best_G), function(k) {
  top <- names(sort(cl_avg_ranks[[k]]))[1:2]
  bot <- names(sort(cl_avg_ranks[[k]], decreasing = TRUE))[1]
  glue::glue("Cluster {k} ({cl_sizes$Count[k]} respondents, {cl_sizes$Proportion[k]*100}%): ",
             "prefers {top[1]} and {top[2]}; likes {bot} least.")
})

# BIC table for reporting (mclustBIC: rows = model names, cols = G values)
bic_vals <- apply(bic_obj, 1, max, na.rm = TRUE)
bic_table <- tibble(
  K     = 1:4,
  BIC   = round(bic_vals, 1)
) |> mutate(`Selected` = if_else(K == best_G, "✓", ""))
```

## Summary of Findings

### Part 1 — Most popular sushi options

All three voting methods point to the same winner:

- **First-preference leader:** `r fp_winner` (`r fp_df$Count[1]` first-place votes, `r round(fp_df$Proportion[1]*100, 1)`% of respondents).
- **IRV winner:** `r irv_winner` (survived all elimination rounds).
- **Condorcet winner:** `r condorcet_label`.

The three methods **agree**: *`r fp_winner`* is unambiguously the most popular sushi option in this dataset. *Cucumber roll* consistently appears at the bottom across all voting rules.

### Part 2 — Agreement among respondents

Kendall's coefficient of concordance is **W = `r round(W, 3)`** (`r W_interp` agreement), which is highly statistically significant (χ²(`r df_W`) = `r round(chi_sq_W, 1)`, *p* < 0.001). Respondents show non-random, consistent preferences for sushi — the agreement is real, not an artefact of chance.

### Part 3 — Preference clusters

A Gaussian mixture model (mclust, model `r best_model`) selected **K = `r best_G` clusters** based on BIC. Each respondent has a posterior probability of belonging to each cluster. Key clusters:

`r paste("- ", cl_desc, collapse = "\n")`

Respondents are assigned **probabilistically** — most belong predominantly to one cluster, but the soft membership captures genuine uncertainty for borderline cases.

---

## Setup and Data

```{r data-peek}
cat("Dimensions:", dim(sushi), "\n")
cat("Items:", paste(items, collapse = ", "), "\n")
head(sushi, 3) |> kable(caption = "First 3 rows of sushi_rankings (1 = most preferred)")
```

The data (`BayesMallows::sushi_rankings`) contains **`r n_voters` respondents** who each ranked **`r n_items` sushi types** from 1 (most preferred) to 10 (least preferred). The sushi types are: `r paste(items, collapse = ", ")`.

---

## Part 1 — Most Popular Options

### First-preference counts

```{r first-pref}
fp_df |>
  mutate(Proportion = scales::percent(Proportion, accuracy = 0.1)) |>
  kable(caption = "First-place vote counts per sushi item (sorted descending)")
```

```{r first-pref-plot, fig.width=7, fig.height=4}
fp_df |>
  mutate(Item = fct_reorder(Item, Count)) |>
  ggplot(aes(x = Item, y = Count,
             fill = Item == fp_winner)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("grey70", "steelblue")) +
  geom_text(aes(label = Count), hjust = -0.2, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_flip() +
  labs(
    title = "First-preference votes per sushi item",
    x = NULL, y = "Number of respondents"
  ) +
  theme_bw() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank())
```

**`r fp_winner`** received the most first-place votes (`r fp_df$Count[1]`, `r round(fp_df$Proportion[1]*100,1)`% of respondents).

### Instant-Runoff Voting (IRV)

```{r irv-table}
irv_round_df <- map_dfr(seq_along(irv_res$rounds), function(i) {
  r <- irv_res$rounds[[i]]
  tibble(Round = i, Item = names(r), Votes = as.integer(r))
})

irv_wide <- irv_round_df |>
  pivot_wider(names_from = Round, values_from = Votes,
              names_prefix = "Round ") |>
  arrange(desc(`Round 1`))

kable(irv_wide, caption = "IRV vote tallies by round (NA = already eliminated)")
```

```{r irv-elim}
if (length(irv_res$eliminated) > 0) {
  cat("Elimination sequence:", paste(irv_res$eliminated, collapse = " → "), "\n")
} else {
  cat("No eliminations needed — a majority was reached in Round 1.\n")
}
cat("IRV winner:", irv_winner)
```

The **IRV winner is `r irv_winner`**, consistent with its dominant first-preference share.

### Condorcet analysis

```{r condorcet-matrix, fig.width=7, fig.height=5.5}
# Show pairwise win matrix as a heatmap
pw_df <- as_tibble(pw_wins, rownames = "Item_A") |>
  pivot_longer(-Item_A, names_to = "Item_B", values_to = "Wins") |>
  filter(Item_A != Item_B) |>
  mutate(Majority = Wins > n_voters / 2)

ggplot(pw_df, aes(x = Item_B, y = Item_A, fill = Wins)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Wins), size = 2.8) +
  scale_fill_gradient2(
    low = "#d73027", mid = "white", high = "#4575b4",
    midpoint = n_voters / 2,
    name = "Voters preferring\nrow over column"
  ) +
  labs(
    title = "Pairwise Condorcet win counts",
    subtitle = paste0("Blue = row item preferred by >", n_voters/2, " voters; red = opposite"),
    x = "Item B", y = "Item A"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r condorcet-result}
if (!is.na(condorcet_winner)) {
  cat("Condorcet winner:", condorcet_winner, "\n")
  cat(condorcet_winner, "beats every other item in head-to-head comparisons.\n")
} else {
  cat("No Condorcet winner exists (preference cycles detected).\n")
  if (!is.null(cycle_example)) {
    cat("Example cycle:", cycle_example[1], "beats", cycle_example[2],
        ",", cycle_example[2], "beats", cycle_example[3],
        ", but", cycle_example[3], "beats", cycle_example[1], "\n")
  }
}
```

**`r condorcet_label`** beats every other sushi item in direct head-to-head pairwise comparisons.

---

## Part 2 — Agreement vs Randomness (Kendall's W)

### Method

Kendall's coefficient of concordance $W$ measures the degree of agreement among $m$ judges ranking $n$ items:

$$W = \frac{12S}{m^2(n^3 - n)}$$

where $S = \sum_{j=1}^{n}(R_j - \bar{R})^2$ is the sum of squared deviations of column rank sums from their expected value under randomness ($\bar{R} = m(n+1)/2$). Under the null hypothesis of random rankings, $m(n-1)W \sim \chi^2_{n-1}$ (large-sample approximation).

```{r kendall-compute}
kendall_tbl <- tibble(
  Statistic = c("m (judges)", "n (items)", "W", "χ² statistic", "df", "p-value"),
  Value     = c(m, n,
                round(W, 4),
                round(chi_sq_W, 2),
                df_W,
                format.pval(p_W, digits = 3, eps = 1e-15))
)
kable(kendall_tbl, caption = "Kendall's W results")
```

```{r kendall-colsums, fig.width=7, fig.height=4}
tibble(Item = items, RankSum = R_j) |>
  mutate(Item = fct_reorder(Item, RankSum)) |>
  ggplot(aes(x = Item, y = RankSum)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = R_bar, linetype = "dashed", colour = "firebrick",
             linewidth = 0.8) +
  annotate("text", x = 1.5, y = R_bar + 500,
           label = paste("Expected under\nrandomness =", R_bar),
           colour = "firebrick", size = 3.2, hjust = 0) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  coord_flip() +
  labs(title = "Column rank sums (deviation from randomness drives W)",
       x = NULL, y = "Sum of ranks across 5 000 respondents") +
  theme_bw() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank())
```

### Interpretation

$W = `r round(W, 3)`$ indicates **`r W_interp` agreement** among the 5 000 respondents. The chi-squared test strongly rejects the null hypothesis of random rankings ($\chi^2(`r df_W`) = `r round(chi_sq_W, 1)`$, $p < 0.001$). While a perfect $W = 1$ is rare for large groups with diverse tastes, the observed level of concordance confirms that sushi preferences are **non-random and consistent across respondents** — fatty tuna and tuna are broadly liked, while cucumber roll and egg are broadly disliked.

---

## Part 3 — Clustering with Probabilistic Class Membership

### Model choice

We use **Gaussian Mixture Models (GMM)** via the `mclust` R package (Scrucca et al., 2023). Although the data are rankings rather than continuous measurements, GMMs applied to rank matrices are a well-established general-purpose latent class approach that directly yields posterior class membership probabilities for each respondent. `mclust` selects the covariance structure automatically via BIC.

> *Justification:* `mclust` handles multivariate continuous (or rank-as-continuous) data and provides exact posterior probabilities $P(\text{cluster } k \mid \text{ranking}_i)$ via the E-step of the EM algorithm — exactly what this analysis requires.

### Number of clusters (BIC)

```{r bic-table}
kable(bic_table, caption = "mclust BIC by number of clusters K (higher = better)")
```

```{r bic-plot, fig.width=6, fig.height=4}
bic_table |>
  ggplot(aes(x = K, y = BIC)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(aes(colour = K == best_G), size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("grey60", "steelblue")) +
  labs(title = "mclust BIC by number of clusters",
       subtitle = paste0("Selected: K = ", best_G, " (", best_model, " model)"),
       x = "Number of clusters K", y = "BIC (higher = better)") +
  theme_bw()
```

`mclust` selected **K = `r best_G`** clusters (model `r best_model`) as the highest-BIC solution.

### Cluster sizes

```{r cluster-sizes}
kable(cl_sizes, caption = paste0("Respondent counts per cluster (K = ", best_G, ")"))
```

### Average ranks per cluster

Lower values indicate stronger preference (rank 1 = most preferred).

```{r cluster-avg-ranks}
avg_rank_df <- map_dfr(seq_len(best_G), function(k) {
  tibble(Cluster = paste0("Cluster ", k),
         Item    = items,
         AvgRank = cl_avg_ranks[[k]])
})

avg_rank_wide <- avg_rank_df |>
  pivot_wider(names_from = Cluster, values_from = AvgRank) |>
  mutate(across(where(is.numeric), ~round(.x, 2))) |>
  arrange(`Cluster 1`)

kable(avg_rank_wide,
      caption = "Average rank per item per cluster (lower = more preferred)")
```

```{r cluster-heatmap, fig.width=7, fig.height=4}
avg_rank_df |>
  mutate(Item = fct_reorder(Item, AvgRank, .fun = mean)) |>
  ggplot(aes(x = Cluster, y = Item, fill = AvgRank)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(AvgRank, 1)), size = 3.5) +
  scale_fill_gradient2(
    low  = "#4575b4",
    mid  = "white",
    high = "#d73027",
    midpoint = (n_items + 1) / 2,
    name = "Avg rank\n(lower = preferred)"
  ) +
  labs(title = "Average preference rank by cluster",
       x = NULL, y = NULL) +
  theme_bw()
```

### Top 3 items per cluster

```{r top3}
tibble(
  Cluster = paste0("Cluster ", seq_len(best_G)),
  `Count (n)` = cl_sizes$Count,
  `Top 3 most preferred` = top3_per_cluster
) |> kable(caption = "Most preferred sushi items per cluster")
```

### Posterior membership probabilities (sample)

The matrix below shows the first 10 respondents' posterior probabilities of belonging to each cluster ($\sum_k P_k = 1$ per row).

```{r posterior-probs}
z_matrix |>
  as_tibble(.name_repair = "unique") |>
  setNames(paste0("P(Cluster ", seq_len(best_G), ")")) |>
  mutate(
    Respondent = seq_len(n_voters),
    `Hard assignment` = cl_assign
  ) |>
  select(Respondent, `Hard assignment`, everything()) |>
  slice_head(n = 10) |>
  mutate(across(starts_with("P("), ~round(.x, 3))) |>
  kable(caption = "Posterior cluster membership probabilities (first 10 respondents)")
```

### Cluster descriptions

```{r cluster-desc}
map_dfr(seq_len(best_G), function(k) {
  avg_r   <- cl_avg_ranks[[k]]
  ord     <- order(avg_r)
  top3    <- names(avg_r)[ord[1:3]]
  bottom2 <- names(avg_r)[ord[(n_items-1):n_items]]
  tibble(
    Cluster    = paste0("Cluster ", k),
    n          = cl_sizes$Count[k],
    `%`        = paste0(round(cl_sizes$Proportion[k]*100, 1), "%"),
    `Top 3 (most preferred)` = paste(top3, collapse = " > "),
    `Bottom 2 (least preferred)` = paste(bottom2, collapse = " > ")
  )
}) |> kable(caption = "Plain-language cluster profiles")
```

---

## Session Info

```{r session-info}
sessioninfo::session_info(c("BayesMallows", "mclust", "tidyverse"))
```
