---
title: "Sushi Rankings Analysis"
subtitle: "Voting rules, agreement, and preference clustering"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
    df-print: kable
execute:
  warning: false
  message: false
  cache: false
---

```{r setup, include=FALSE}
library(prefviz)
library(prefio)
library(tidyverse)
library(mclust)
library(knitr)

# ── Load sushi data from PrefLib ──────────────────────────────────────────────
sushi_data <- prefio::read_preflib(
  "00014 - sushi/00014-00000001.soc",
  from_preflib = TRUE
)

# ── Clustering (mclust on PCA) ──────────────────────────────────────
items     <- levels(sushi_data$preferences)
n_voters  <- sum(sushi_data$frequency) 
n_items   <- length(items)               # 10

# Expand to one row per respondent for matrix-based analyses
sushi_prefs_expanded <- rep(sushi_data$preferences, sushi_data$frequency)
rankings_mat <- prefio::ranking_matrix(sushi_prefs_expanded)
colnames(rankings_mat) <- items

# Rankings sum to 55 per row → data lies on a 9-D hyperplane.
# Project onto first 9 PCs before fitting GMM to avoid singular covariances.
pca_sushi  <- prcomp(rankings_mat, center = TRUE, scale. = FALSE)
sushi_9pc  <- pca_sushi$x[, 1:9]

set.seed(42)
bic_obj  <- mclustBIC(sushi_9pc, G = 1:4)
mc_fit   <- Mclust(sushi_9pc, G = 1:4, x = bic_obj)
best_G   <- mc_fit$G
best_model <- mc_fit$modelName

z_matrix  <- mc_fit$z          # posterior membership probabilities
cl_assign <- mc_fit$classification

cl_sizes <- tibble(
  Cluster    = seq_len(best_G),
  Count      = as.integer(table(cl_assign)),
  Proportion = round(as.numeric(table(cl_assign)) / n_voters, 3)
)

cl_avg_ranks <- lapply(seq_len(best_G), function(k) {
  colMeans(rankings_mat[cl_assign == k, ])
})

top3_per_cluster <- map_chr(cl_avg_ranks, function(avg_r) {
  paste(names(sort(avg_r))[1:3], collapse = ", ")
})

cl_desc <- map_chr(seq_len(best_G), function(k) {
  top <- names(sort(cl_avg_ranks[[k]]))[1:2]
  bot <- names(sort(cl_avg_ranks[[k]], decreasing = TRUE))[1]
  glue::glue(
    "Cluster {k} ({cl_sizes$Count[k]} respondents, ",
    "{cl_sizes$Proportion[k] * 100}%): ",
    "prefers {top[1]} and {top[2]}; likes {bot} least."
  )
})

bic_vals  <- apply(bic_obj, 1, max, na.rm = TRUE)
bic_table <- tibble(
  K        = 1:4,
  BIC      = round(bic_vals, 1),
  Selected = if_else(K == best_G, "✓", "")
)

# ── Part 3: Ternary plot of cluster membership (K = 3) ───────────────────────
# Posterior probs sum to 1 per respondent → natural simplex coordinates.
# Use the optimal K if it equals 3, otherwise fit a dedicated K=3 model.
if (best_G == 3) {
  z_tern     <- z_matrix
  cl_tern    <- cl_assign
} else {
  mc_fit3  <- Mclust(sushi_9pc, G = 3)
  z_tern   <- mc_fit3$z
  cl_tern  <- mc_fit3$classification
}

cl_names <- paste0("Cluster ", seq_len(3))
tern_probs_df <- as.data.frame(z_tern)
colnames(tern_probs_df) <- cl_names
tern_probs_df$assignment <- paste0("Cluster ", cl_tern)

tern_obj  <- as_ternable(tern_probs_df, items = all_of(cl_names))
tern_data <- get_tern_data2d(tern_obj)
```

## Setup and Data

```{r data-peek}
sushi_data <- prefio::read_preflib(
  "00014 - sushi/00014-00000001.soc",
  from_preflib = TRUE
)
sushi_data
```

The data are loaded via `prefio::read_preflib()` from the PrefLib sushi dataset
(Kato & Ishigami 2004; file `00014-00000001.soc`). Each of the **`r n_voters`
respondents** gave a complete ranking of **`r n_items` sushi types** from 1
(most preferred) to 10 (least preferred). The sushi types are:
`r paste(items, collapse = ", ")`.

---

## Part 1 — Most Popular Options

### First-preference counts

```{r first-pref-table}
irv_tbl <- dop_irv(
  sushi_data,
  value_type    = "percent",
  preferences_col = preferences,
  frequency_col   = frequency
)
irv_winner <- unique(irv_tbl$winner)

dop_bar(irv_tbl, -c(round, winner), at_round = 1)
```

**`r irv_winner`** received the most first-place votes

### Instant-Runoff Voting (IRV)

IRV results are computed using `prefviz::dop_irv()`.

```{r irv-table}
irv_wide <- irv_tbl |>
  select(-winner) |>
  rename(Round = round) |>
  arrange(Round)

kable(irv_wide,
      caption = "IRV vote tallies by round (NA = already eliminated)")
```

The **IRV winner is `r irv_winner`**, consistent with its dominant
first-preference share.

### Condorcet analysis

Pairwise comparisons are computed using `prefviz::pairwise_calculator()`

```{r condorcet-heatmap, fig.width=7, fig.height=5.5}
pw_result <- pairwise_calculator(
  sushi_data,
  preferences_col = preferences,
  frequency_col   = frequency
)
pw_result
```

```{r}
pairwise_heatmap(pw_result, value = "tcp")
```

## Part 2 — Clustering with Probabilistic Class Membership

### Model choice

We use **Gaussian Mixture Models (GMM)** via the `mclust` R package
(Scrucca et al., 2023). Although the data are rankings rather than continuous
measurements, GMMs applied to rank matrices are a well-established
general-purpose latent class approach that directly yields posterior class
membership probabilities for each respondent. `mclust` selects the covariance
structure automatically via BIC.

> *Justification:* `mclust` handles multivariate continuous (or
> rank-as-continuous) data and provides exact posterior probabilities
> $P(\text{cluster } k \mid \text{ranking}_i)$ via the E-step of the EM
> algorithm — exactly what this analysis requires. The ranking matrix is
> first projected onto 9 principal components to remove the zero-variance
> direction (all row sums are identical = 55).

### Number of clusters (BIC)

```{r bic-table}
kable(bic_table,
      caption = "mclust BIC by number of clusters K (higher = better)")
```

```{r bic-plot, fig.width=6, fig.height=4}
bic_table |>
  ggplot(aes(x = K, y = BIC)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(aes(colour = K == best_G), size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("grey60", "steelblue")) +
  labs(
    title = "mclust BIC by number of clusters",
    subtitle = paste0("Selected: K = ", best_G, " (", best_model, " model)"),
    x = "Number of clusters K", y = "BIC (higher = better)"
  ) +
  theme_bw()
```

`mclust` selected **K = `r best_G`** clusters (model `r best_model`) as the
highest-BIC solution.

### Cluster sizes

```{r cluster-sizes}
kable(cl_sizes,
      caption = paste0("Respondent counts per cluster (K = ", best_G, ")"))
```

### Average ranks per cluster

Lower values indicate stronger preference (rank 1 = most preferred).

```{r cluster-avg-ranks}
avg_rank_df <- map_dfr(seq_len(best_G), function(k) {
  tibble(
    Cluster = paste0("Cluster ", k),
    Item    = items,
    AvgRank = cl_avg_ranks[[k]]
  )
})

avg_rank_df |>
  pivot_wider(names_from = Cluster, values_from = AvgRank) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  arrange(`Cluster 1`) |>
  kable(caption = "Average rank per item per cluster (lower = more preferred)")
```

```{r cluster-heatmap, fig.width=7, fig.height=4}
avg_rank_df |>
  mutate(Item = fct_reorder(Item, AvgRank, .fun = mean)) |>
  ggplot(aes(x = Cluster, y = Item, fill = AvgRank)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(AvgRank, 1)), size = 3.5) +
  scale_fill_gradient2(
    low      = "#4575b4",
    mid      = "white",
    high     = "#d73027",
    midpoint = (n_items + 1) / 2,
    name     = "Avg rank\n(lower = preferred)"
  ) +
  labs(title = "Average preference rank by cluster", x = NULL, y = NULL) +
  theme_bw()
```

### Top 3 items per cluster

```{r top3}
tibble(
  Cluster              = paste0("Cluster ", seq_len(best_G)),
  `Count (n)`          = cl_sizes$Count,
  `Top 3 most preferred` = top3_per_cluster
) |> kable(caption = "Most preferred sushi items per cluster")
```

### Posterior membership probabilities (sample)

The matrix below shows the first 10 respondents' posterior probabilities of
belonging to each cluster ($\sum_k P_k = 1$ per row).

```{r posterior-probs}
z_matrix |>
  as_tibble(.name_repair = "unique") |>
  setNames(paste0("P(Cluster ", seq_len(best_G), ")")) |>
  mutate(
    Respondent       = seq_len(n_voters),
    `Hard assignment` = cl_assign
  ) |>
  select(Respondent, `Hard assignment`, everything()) |>
  slice_head(n = 10) |>
  mutate(across(starts_with("P("), ~ round(.x, 3))) |>
  kable(caption = "Posterior cluster membership probabilities (first 10 respondents)")
```

### Ternary plot of cluster membership

The 2D ternary plot below uses the posterior probabilities from a K = 3 model
as simplex coordinates (each respondent's three cluster probabilities sum to 1,
placing them inside the triangle). Points near a vertex have high certainty of
belonging to that cluster; points near the centre are uncertain across all three.
Built with `prefviz::as_ternable()` and `prefviz::get_tern_data2d()`.

```{r ternary-cluster, fig.width=6, fig.height=5.5}
cluster_colours <- c(
  "Cluster 1" = "#E13940",
  "Cluster 2" = "#1C4F9C",
  "Cluster 3" = "#10C25B"
)

ggplot(tern_data, aes(x = x1, y = x2)) +
  add_ternary_base() +
  geom_point(
    aes(colour = assignment),
    alpha = 0.25, size = 0.8
  ) +
  add_vertex_labels(tern_obj$simplex_vertices, size = 4, fontface = "bold") +
  scale_colour_manual(values = cluster_colours, name = "Hard assignment") +
  labs(title = "Ternary plot of cluster membership probabilities (K = 3)") +
  guides(colour = guide_legend(
    override.aes = list(alpha = 1, size = 3)
  ))
```

### Cluster descriptions

```{r cluster-desc}
map_dfr(seq_len(best_G), function(k) {
  avg_r   <- cl_avg_ranks[[k]]
  ord     <- order(avg_r)
  top3    <- names(avg_r)[ord[1:3]]
  bottom2 <- names(avg_r)[ord[(n_items - 1):n_items]]
  tibble(
    Cluster                    = paste0("Cluster ", k),
    n                          = cl_sizes$Count[k],
    `%`                        = paste0(round(cl_sizes$Proportion[k] * 100, 1), "%"),
    `Top 3 (most preferred)`   = paste(top3, collapse = " > "),
    `Bottom 2 (least preferred)` = paste(bottom2, collapse = " > ")
  )
}) |> kable(caption = "Plain-language cluster profiles")
```