---
title: "Using `ternable` object to draw ternary plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using `ternable` object to draw ternary plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

```{r setup}
#| echo: false
library(prefviz)
library(ggplot2)
library(tourr)
library(dplyr)
```

This vignette shows you how to build a ternary plot on 2 and higher dimensions, using the `ternable` object.

Both 2D and high-dimensional (HD) ternary plots require the following 3 components:

- Coordinates of the observations: Your n-part compositional data must be transformed into (n-1)-dimensional space via Helmert matrix.
- Vertices: The point coordinates that define the vertices of the simplex
- Edges: How the vertices are connected to create the simplex

You can access all these components conveniently via a `ternable` object.

# `ternable` object

`ternable` is a simple S3 object that contains all the data and metadata useful for ternary plots, including the following components:

- `data`: Stores input data after being validated and normalized 
- `data_coord`: Stores the coordinates for all observations.
- `data_edges`: Stores the connections between the observations. Useful when you want to create paths between observations. 
- `simplex_vertices`: Stores the simplex vertices' coordinates.
- `simplex_edges`: Stores the connections between the simplex vertices.
- `vertex_labels`: Stores the vertex labels/item names in the order provided in the argument `items`.

To create a `ternable` object, simply call the function `ternable()`. `ternable()` takes 2 arguments:

- `data`: The input data, which must be in a `ternable`-friendly format. For more details on how to transform your raw data into a `ternable`-friendly format, please refer to `vignette("transform_raw_data")`
- `items`: The item names in the order you want them to appear in the ternary plot. The default takes all the columns in `data`.

```{r}
aecdop22_transformed <- prefviz:::aecdop22_transformed
head(aecdop22_transformed)

tern22 <- ternable(data = aecdop22_transformed, items = ALP:Other)
tern22
```

## `ternable` helpers - `get_tern_*()`

While `ternable` provides you with the essenstial components for building a ternary plot, different plot types (2D or HD) might require slightly different way of representing these commponents. `get_tern_*()` functions help you do just that. 

Under the hood, `get_tern_*()` perform simple data transformations, i.e., `rbind()` and `cbind()`, to help you create the input that are compatible with popular plotting packages, i.e., `ggplot2` for 2D ternary plot and `tourr` for HD ternary plots.

There are 3 `get_tern_*()` functions:

- `get_tern_data()`: Provides input data for `ggplot2` or `tourr`.
- `get_tern_edges()`: Provides edges of the simplex, which is required by `tourr`. 
- `get_tern_labels()`: Provides labels for the vertices, which is helpful for distinguishing vertices in `tourr`.

# Drawing a 2D ternary plot

Take the example of the 2022 Australian Federal Election, we would like to take a look at the first preference distribution between the 2 major parties: Labor and the Coalition, and other parties. 

The dataset `aecdop22_transformed` is already in a `ternable`-friendly format, so we can directly pass it to `ternable()` to create a `ternable` object.

```{r}
tern22 <- ternable(aecdop22_transformed, ALP:Other)
```

Now we can use the `get_tern_data()` function to get the input data for `ggplot2`.

```{r}
input_df <- get_tern_data(tern22, plot_type = "2D")
head(input_df)
```

The output is a data frame where the original columns are combined with the coordinates (`x1`, `x2`). These coordinate columns are the observation locations on the plot. We can now use `ggplot2` to draw the ternary plot.

```{r}
p <- ggplot(input_df, aes(x = x1, y = x2)) +
  # Draw the ternary space as an equilateral triangle
  geom_ternary_cart() + 
  # Plot the observations as points
  geom_point(aes(color = ElectedParty)) + 
  # Add vertex labels, taken from the ternable object
  add_vertex_labels(tern22$simplex_vertices) + 
  labs(title = "First preference in 2022 Australian Federal election")

p
```

In an election, we would be interested in defining the regions where one party takes the majority over others. We can do that using `geom_ternary_region()`. 

This geom takes the barycentric coordinates of a reference point as input, and divides the ternary triangle into 3 regions based on the reference points. These regions are defined by the perpendicular projections of the reference point to the three edges of the triangle. The default reference point is the centroid, which divides the triangle into 3 equal regions.

```{r}
p + 
  geom_ternary_region(
    x1 = 1/3, x2 = 1/3, x3 = 1/3, # Default reference points. Must sum to 1
    vertex_labels = tern22$vertex_labels, # Labels for the regions
    aes(fill = after_stat(vertex_labels)), 
    alpha = 0.3, color = NA, show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c("ALP" = "red", "LNP" = "blue", "Other" = "grey70"),
    aesthetics = c("fill", "colour")
  )
```

`vertex_labels` argument is used to specify the vertex of which the region belongs to. This is helpful when you want to "sync" the aesthetic mapping of `geom_ternary_region()` with the base layer because you only need to specify the customization once. 

Please note that the order in which the labels are provided must match the order of the vertices in the ternary plot. The vertices are listed clockwise, from the right (ALP) to the left (LNP), then ending at the top of the triangle (Other). The best way is to get these labels from `ternable$vertex_labels` as `ternable` preserves the vertex orders.

# Drawing a high-dimensional ternary plot

Take the example of the 2025 Australian Federal Election, we would like to take a look at the first preference distribution between the 4 major groups: Labor, Coalition, Greens, Independents and the other party. This can be conveniently done using the `tourr` package, `ternable` object and the `get_tern_*()` functions.

A ternary tour requires the following components:

- Coordinates of the observations and vertices
- Edges of the simplex
- Vertex labels (good to have to identify the vertices during the tour)

```{r}
# Load the data
aecdop25_transformed <- prefviz:::aecdop25_transformed
head(aecdop25_transformed)

tern25 <- ternable(aecdop25_transformed, ALP:IND)

# Animate the tour
animate_xy(
  get_tern_data(tern25, plot_type = "HD"), # Dataframe with coordinates of the observations and vertices
  edges = get_tern_edges(tern25), # Edges of the simplex
  obs_labels  = get_tern_labels(tern25), # Labels for the vertices
  axes = "bottomleft"
)
```

We can add colors to the points.
```{r}
# Define color mapping
party_colors <- c(
  "ALP" = "#E13940",    # Red
  "LNP" = "#1C4F9C",    # Blue
  "GRN" = "#10C25B",    # Green
  "IND" = "#F39C12",    # Orange
  "Other" = "#95A5A6"   # Gray
)

# Map to your data (assuming your column is called elected_party)
color_vector <- c(rep("black", 5),
  party_colors[aecdop25_transformed$ElectedParty])

# Animate the tour
animate_xy(
  get_tern_data(tern25, plot_type = "HD"), 
  edges = get_tern_edges(tern25),
  obs_labels  = get_tern_labels(tern25),
  col = color_vector,
  axes = "bottomleft"
)
```